# Reproducible Research: Peer Assessment 1

July 2014 - Javier González Hernández


## Loading and preprocessing the data

Environment setup.
```{r environmentSetup}
setwd("C:/Users/Jg/RepData_PeerAssessment1")
library(ggplot2)
library(lattice)
```

Loading data and first view.
```{r loadingData}
activity <- read.csv("activity.csv", colClasses=c("numeric", "character", "numeric"), header=T)
head(activity)
```

Extraction of times from intervals.
```{r timeExtraction}
# returns a character vector containing times formatted with 4 digits.
intervals_to_times <- sprintf("%04d", activity$interval)
# replacement of time column with "dd:dd" pattern. Convertion to factor too.
activity$time <- as.factor(gsub("(\\d{2})(\\d{2})","\\1:\\2", intervals_to_times))
head(activity)
```

Preprocesing the data and first approaching plot.
```{r 1stApproach}
# add all steps of each day.
totalStepsByDay <- aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)
head(totalStepsByDay)
# plot with bars for better understanding of data.
days <- totalStepsByDay[,1]
steps <- totalStepsByDay[,2]
ggplot(totalStepsByDay) + 
  geom_bar(aes(x=days, y=steps), fill="yellow",color="black", binwidth=5000, stat="identity") +
  coord_flip()
```


## What is mean total number of steps taken per day?

```{r histogram}
hist(totalStepsByDay$steps, main="Total number of steps taken per day", xlab="steps",
     col="yellow")
```

Mean and median total number of steps taken per day.
```{r meanAndMedian}
mean(totalStepsByDay$steps)
median(totalStepsByDay$steps)
```

Second plot showing the reported mean.
```{r histogramWithMean}
# for better clarity
meanStepsByDay <- mean(totalStepsByDay$steps)
ggplot(totalStepsByDay, aes(x=steps)) +  
  geom_histogram(colour="black", fill="yellow", binwidth=5000) +  
  geom_vline(aes(xintercept=meanStepsByDay, color="Mean"), size=1, show_guide=T) +  
  guides(colour=guide_legend(title=NULL)) +
  labs(title="Total number of steps taken per day showing the mean", x="Number of steps", y="Frequency")
```

## What is the average daily activity pattern?

Preparing and viewing the data.
```{r averageDailyActivity}
# steps by times.The mean.
meanStepsByInterval <- aggregate(activity$steps ~ activity$time, data=activity, mean,
                                 na.rm=TRUE)
head(meanStepsByInterval)
tail(meanStepsByInterval)
```

Ploting the data.
```{r averagePlot}
averageIntervals <- meanStepsByInterval[,1]
averageSteps <- meanStepsByInterval[,2]
ggplot(meanStepsByInterval) +  
  geom_line(aes(y=averageSteps, x=averageIntervals, group=1)) +
  scale_x_discrete(breaks=averageIntervals[seq(1,288, by=6)]) +
  theme(axis.text.x=element_text(angle=90, hjust=1)) + coord_flip() +
  labs(title="Average daily activity pattern", x="average intervals of 5 minutes",
       y="Average number of steps")
```

Interval of 5 minutes that contains the maximum average number of steps.

Maximum average number of steps.
```{r 5minutesMaximun}
maxSteps <- max(meanStepsByInterval[,2])
c(Maximun_average_number_steps=maxSteps)
```

Maximum Interval.
```{r 5minutesInterval}
maxActivityId <- which(meanStepsByInterval[,2]==maxSteps)
c(Maximun_interval=maxActivityId)
```

Interval content.
```{r 5minutesContent}
c(Interval_content=meanStepsByInterval[maxActivityId,])
```

## Imputing missing values

Calculating and reporting the total number of missing values in the dataset.
Only missings values located in steps column.
```{r seekingNAs}
c(MissingValues_Total=sum(is.na(activity)), 
  MissingValues_InSteps=sum(is.na(activity$steps)))
```

Filling the missing values with the 5 minutes interval mean.
```{r fillingNAs}
# steps by intervals.The mean.
meanSteps <- aggregate(steps ~ interval, data=activity, mean)
# container for recalculated steps.
filledNAs <- numeric()
for (i in 1:nrow(activity)) {
    item <- activity[i, ]
    # when missing value replace with the interval mean.
    if (is.na(item$steps)) {
        steps <- subset(meanSteps, interval==item$interval)$steps
    } else {
        steps <- item$steps
    }
    filledNAs <- c(filledNAs, steps)
}
```

Creating the new Dataset without missing values.
```{r newDataset}
newActivity <- activity
# replace with recalculated steps.
newActivity$steps <- filledNAs
head(newActivity)
tail(newActivity)
```

Histogram with the total number of steps taken each day without missings values.
```{r histogramWithoutNAs}
# add all steps of each day. Recalculated steps.
totalSteps <- aggregate(steps ~ date, data=newActivity, sum, na.rm=TRUE)
hist(totalSteps$steps, main="total number of steps taken each day", xlab="steps",
     col="yellow")
```

Calculation and information of the mean and median total number of steps taken per day.
```{r}
c(New_Mean_is=mean(totalSteps$steps))
c(New_Median_is=median(totalSteps$steps))
```

The impact of imputing missing data on the estimates of the total daily number of steps is that
the mean and the median are equal now.

## Are there differences in activity patterns between weekdays and weekends?

Preparing the data.
```{r differencesInActivity}
# converts date column to Date type with format.
activity$date <- as.Date(activity$date, "%Y-%m-%d")
# extract names of the week days.
day <- weekdays(activity$date)
head(day)
```
Comment, my system has Spanish language.

This is a simple iteration to percolate weekdays from weekends
```{r percolatingWeekends}
# determines for each day whether is a weekend day or not.
daytype <- vector()
for (i in 1:nrow(activity)) {
    if (day[i]=="sábado") {
        daytype[i] <- "Weekend"
    } else if (day[i]=="domingo") {
        daytype[i] <- "Weekend"
    } else {
        daytype[i] <- "Weekday"
    }
}
```

Adding and factoring daytype column to data.
```{r factoringDaylevel}
activity$daytype <- daytype
activity$daytype <- factor(activity$daytype)
```

Plotting differences in activity patterns between weekdays and weekends.
```{r differencesPlot}
# groups the data by the day type.
meanInterval <- aggregate(steps ~ interval + daytype, data=activity, mean)
xyplot(steps ~ interval | daytype, meanInterval, type="l", layout=c(1, 2), 
    xlab="Interval", ylab="Number of steps")
```

The plot shows differences.

## Resources and Bibliography used in this assessment.

[Johns Hopkins Bloomberg School of Public Health Coursera course - Reproducible Research](https://www.coursera.org/course/repdata/)

[Johns Hopkins Bloomberg School of Public Health Coursera course - Exploratory Data Analysis](https://www.coursera.org/course/exdata)

[The R Manuals - CRAN site](http://cran.r-project.org/)

[R in Action, Second Edition](http://www.manning.com/kabacoff2/)

[R Graphics Cookbook](http://shop.oreilly.com/product/0636920023135.do)

[Practical Data Science with R](http://www.manning.com/zumel/)

[R For Dummies, 2nd Edition](http://www.dummies.com/store/product/R-For-Dummies.productCd-1119962846,navId-322468,descCd-tableOfContents.html)

[Every book in this assessment used with the safari books online service.](https://www.safaribooksonline.com/) 
